---
title: "The basic reproduction number: threshold and intuition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The basic reproduction number: threshold and intuition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(epiLabR)
```

## What is R₀?

The **basic reproduction number**, denoted $R_0$, is perhaps the most
important concept in mathematical epidemiology. Its definition is straightforward:

> $R_0$ is the average number of secondary cases generated by a single
> infectious individual over their entire infectious period, in a **fully
> susceptible** population.

In the classic SIR model, each infectious individual:

- Transmits the pathogen at rate $\beta$ (contacts per unit time
  multiplied by the probability of transmission per contact).
- Remains infectious for a mean duration of $1/\gamma$ days.

Therefore, the expected number of new infections it produces is:

$$R_0 = \frac{\beta}{\gamma}$$

**Intuitive example.** If $\beta = 0{,}3$ day$^{-1}$ and $\gamma = 0{,}1$
day$^{-1}$, the mean infectious period is 10 days and each infected individual
generates on average $0{,}3 \times 10 = 3$ secondary cases. The epidemic grows
rapidly.

---

## The threshold theorem

The fundamental result of Kermack and McKendrick (1927) states:

$$\text{Epidemic} \iff R_0 > 1$$

- If $R_0 \leq 1$: each infectious individual generates at most one new case. The
  outbreak goes extinct without producing an appreciable epidemic.
- If $R_0 > 1$: each infectious individual generates more than one new case. The
  infection spreads exponentially at first and culminates in an epidemic.

The logic is direct: when $R_0 > 1$ the rate of new infections exceeds the
recovery rate, so $I(t)$ increases. When $R_0 < 1$ the opposite occurs and
$I(t)$ decreases from the very start.

---

## Simulation: three scenarios

We will observe the threshold theorem in action by simulating the SIR model
with three values of $R_0$.

```{r parametros}
N     <- 1e6   # population size
I0    <- 10    # initial infectious individuals
gamma <- 0.1   # recovery rate (mean infectious period = 10 days)
times <- 0:300 # time horizon (days)

escenarios <- list(
  list(R0 = 0.8, beta = 0.08),
  list(R0 = 1.5, beta = 0.15),
  list(R0 = 3.0, beta = 0.30)
)
```

```{r simulaciones, echo=TRUE}
sims <- lapply(escenarios, function(e) {
  simulate_epi(
    model = SIR_MODEL,
    times = times,
    parms = c(beta = e$beta, gamma = gamma),
    init  = c(S = N - I0, I = I0, R = 0)
  )
})
```

```{r grafico-I, fig.width=7, fig.height=4.5, fig.cap="Epidemic curves I(t) for three values of R0. When R0 < 1 the outbreak goes extinct; when R0 > 1 an epidemic occurs, growing in size with R0."}
colores   <- c("steelblue", "darkorange", "firebrick")
etiquetas <- c(expression(R[0] == 0.8),
               expression(R[0] == 1.5),
               expression(R[0] == 3.0))

ylim_max <- max(sapply(sims, function(s) max(s$states$I)))

plot(
  times, sims[[1]]$states$I,
  type = "l", lwd = 2, col = colores[1],
  ylim = c(0, ylim_max * 1.05),
  xlab = "Time (days)",
  ylab = "Infectious individuals I(t)",
  main = expression("Epidemic curves for different " * R[0])
)
lines(times, sims[[2]]$states$I, lwd = 2, col = colores[2])
lines(times, sims[[3]]$states$I, lwd = 2, col = colores[3])
abline(h = 0, lty = 3, col = "gray50")
legend(
  "topright",
  legend  = etiquetas,
  col     = colores,
  lwd     = 2,
  bty     = "n"
)
```

We observe that:

- **$R_0 = 0{,}8$** (blue): infectious individuals decrease monotonically from
  day one. No epidemic occurs.
- **$R_0 = 1{,}5$** (orange): a moderate outbreak occurs with a clearly
  visible peak.
- **$R_0 = 3{,}0$** (red): the epidemic is much more intense and reaches its
  peak earlier.

---

## Final epidemic size

When the epidemic ends, what fraction of the population will have been
infected? The **final size relation** states that the fraction $z$
of infected individuals at the end satisfies the transcendental equation:

$$z = 1 - e^{-R_0 \, z}$$

It has no closed-form analytical solution, but we can compute it numerically
from the simulations as $z = 1 - S(\infty)/N$.

```{r tamano-final, echo=TRUE}
tamano_final <- sapply(sims, function(s) {
  S_final <- tail(s$states$S, 1)
  1 - S_final / N
})

resultado <- data.frame(
  R0            = sapply(escenarios, function(e) e$R0),
  beta          = sapply(escenarios, function(e) e$beta),
  gamma         = gamma,
  tamano_final  = round(tamano_final, 4)
)

print(resultado)
```

```{r grafico-tamano, fig.width=5, fig.height=4, fig.cap="Final fraction of the population infected as a function of R0."}
barplot(
  resultado$tamano_final,
  names.arg = paste0("R0 = ", resultado$R0),
  col       = colores,
  ylab      = "Infected fraction",
  main      = "Final epidemic size",
  ylim      = c(0, 1),
  border    = NA
)
abline(h = 0, col = "gray50")
```

As expected, the higher $R_0$, the larger the fraction of the population
that gets infected. With $R_0 = 0{,}8$ practically no additional individuals
are infected; with $R_0 = 3{,}0$ the vast majority of the population passes
through the infectious compartment.

---

## The effective reproduction number $R_t$

As the epidemic progresses, the fraction of susceptibles $S(t)/N$
decreases because infected individuals recover and acquire immunity. This means
the pathogen finds fewer and fewer available targets. The **effective
reproduction number** captures this dynamic:

$$R_t = R_0 \cdot \frac{S(t)}{N}$$

When $R_t$ crosses the value 1 downward, the epidemic reaches its **peak**:
from that moment on there are more recoveries than new infections and $I(t)$
begins to decline.

```{r Rt-pico, fig.width=7, fig.height=4.5, fig.cap="Rt = R0 * S(t)/N for the R0 = 3 scenario. The dotted line marks Rt = 1; the crossing coincides with the peak of I(t)."}
# Use the R0 = 3.0 scenario (third element)
s3   <- sims[[3]]
R0_3 <- escenarios[[3]]$R0

Rt <- R0_3 * s3$states$S / N

# Moment of the I(t) peak
t_pico <- times[which.max(s3$states$I)]

par(mar = c(4, 4, 3, 4))
plot(
  times, Rt,
  type = "l", lwd = 2, col = "firebrick",
  xlab = "Time (days)",
  ylab = expression(R[t]),
  main = expression(R[t] * " and epidemic peak  (" * R[0] == 3.0 * ")")
)
abline(h = 1,       lty = 2, col = "gray40")
abline(v = t_pico,  lty = 3, col = "gray40")
text(t_pico + 5, max(Rt) * 0.9,
     paste0("peak: day ", t_pico),
     adj = 0, cex = 0.85)
legend("topright",
       legend = c(expression(R[t]), expression(R[t] == 1), "peak I(t)"),
       col    = c("firebrick", "gray40", "gray40"),
       lty    = c(1, 2, 3),
       lwd    = c(2, 1, 1),
       bty    = "n")
```

**Conclusion:** the criterion $R_t = 1$ is dynamic. Although at the start of the
epidemic $R_t \approx R_0 > 1$, as the susceptible population is depleted
$R_t$ declines. The epidemic is controlled ---naturally, without
intervention--- when $R_t$ falls below 1. Public health interventions
(vaccination, isolation, distancing) aim to reduce $R_t$ below that threshold
as quickly as possible.

---

## Summary

| Concept | Formula | Interpretation |
|---|---|---|
| Basic reproduction number | $R_0 = \beta/\gamma$ | Secondary cases in a fully susceptible population |
| Epidemic threshold | $R_0 > 1$ | Necessary and sufficient condition for an epidemic |
| Final size | $z = 1 - S(\infty)/N$ | Fraction of the population infected at the end of the outbreak |
| Effective reproduction | $R_t = R_0 \cdot S(t)/N$ | Actual transmissibility at each moment; peak when $R_t = 1$ |

---

## References

Kermack, W. O. & McKendrick, A. G. (1927). A contribution to the mathematical
theory of epidemics. *Proceedings of the Royal Society of London A*,
**115**(772), 700--721. <https://doi.org/10.1098/rspa.1927.0118>

Anderson, R. M. & May, R. M. (1991). *Infectious Diseases of Humans: Dynamics
and Control*. Oxford University Press.

Hethcote, H. W. (2000). The mathematics of infectious diseases. *SIAM Review*,
**42**(4), 599--653. <https://doi.org/10.1137/S0036144500371907>
