---
title: "The latent period: SIR versus SEIR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The latent period: SIR versus SEIR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(epiLabR)
```

## Motivation: the latent period

Many infectious diseases present a **latent period**: the time that elapses between
exposure to the pathogen and the onset of infectiousness.
During this period the individual is already infected ---and therefore no longer
susceptible--- but cannot yet transmit the infection.

Some representative examples:

| Disease       | Mean latent period |
|---------------|--------------------|
| COVID-19      | ~5 days            |
| Influenza     | ~2 days            |
| Measles       | ~10 days           |

The **SIR** model ignores this phase and assumes that exposure and infectiousness
are simultaneous.
The **SEIR** model represents it explicitly through an $E$ (*Exposed*) compartment
and the parameter $\sigma$: the progression rate from $E$ to $I$.
The mean latent period is $1/\sigma$ days.

## The SEIR model

The system of differential equations for the SEIR is:

$$
\frac{dS}{dt} = -\lambda(t), \quad
\frac{dE}{dt} = \lambda(t) - \sigma E, \quad
\frac{dI}{dt} = \sigma E - \gamma I, \quad
\frac{dR}{dt} = \gamma I,
$$

where $\lambda(t) = \beta \, S(t) \, I(t) / N$ is the force of infection.

## Key insight: $R_0$ does not depend on $\sigma$

The basic reproduction number remains:

$$R_0 = \frac{\beta}{\gamma}$$

The latent period **does not alter the invasion threshold**: if $R_0 > 1$ the
epidemic takes off regardless of $\sigma$.
What does change is the *initial growth rate* and, therefore, **the peak arrives
later** in the SEIR than in the SIR with identical $\beta$ and $\gamma$.

## Comparison: SIR versus SEIR

We use $\beta = 0.3$ and $\gamma = 0.14$, giving $R_0 \approx 2.1$ in both models.
For the SEIR we add $\sigma = 0.2$ (latent period of 5 days).

```{r sir-vs-seir}
parms_sir  <- c(beta = 0.3, gamma = 0.14)
parms_seir <- c(beta = 0.3, sigma = 0.2, gamma = 0.14)
init_sir   <- c(S = 999990, I = 10, R = 0)
init_seir  <- c(S = 999990, E = 0, I = 10, R = 0)
tiempos    <- seq(0, 300, by = 0.5)

sim_sir  <- simulate_epi(SIR_MODEL,  times = tiempos, parms = parms_sir,  init = init_sir)
sim_seir <- simulate_epi(SEIR_MODEL, times = tiempos, parms = parms_seir, init = init_seir)

I_sir  <- sim_sir$states$I
I_seir <- sim_seir$states$I
t      <- sim_sir$states$time

ylim_max <- max(I_sir, I_seir) * 1.05

plot(t, I_sir,
     type = "l", lwd = 2, col = "steelblue",
     ylim = c(0, ylim_max),
     xlab = "Time (days)", ylab = "Infected I(t)",
     main = "SIR vs SEIR: same R0, different peak timing")
lines(t, I_seir, lwd = 2, col = "firebrick", lty = 2)
legend("topright",
       legend = c("SIR", "SEIR (sigma = 0.2, latency 5 days)"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), lwd = 2, bty = "n")
```

The peak of infected individuals is nearly identical in height (same final epidemic
size), but the SEIR reaches that peak considerably later.

## Effect of the latent period: varying $\sigma$

The longer the latent period (smaller $\sigma$), the further the peak shifts to
the right and the slightly flatter it becomes.

```{r sigma-effect}
sigmas     <- c(0.5, 0.2, 0.05)
etiquetas  <- c("sigma = 0.5 (2 days)", "sigma = 0.2 (5 days)", "sigma = 0.05 (20 days)")
colores    <- c("darkorange", "firebrick", "purple")
ltipos     <- c(1, 2, 3)

# Simulate the three scenarios
sims_sigma <- lapply(sigmas, function(s) {
  simulate_epi(SEIR_MODEL,
               times = tiempos,
               parms = c(beta = 0.3, sigma = s, gamma = 0.14),
               init  = init_seir)
})

I_vals <- lapply(sims_sigma, function(s) s$states$I)
ylim_s <- max(unlist(I_vals)) * 1.05

plot(tiempos, I_vals[[1]],
     type = "l", lwd = 2, col = colores[1], lty = ltipos[1],
     ylim = c(0, ylim_s),
     xlab = "Time (days)", ylab = "Infected I(t)",
     main = "SEIR: effect of the latent period (1/sigma)")
for (k in 2:3) {
  lines(tiempos, I_vals[[k]], lwd = 2, col = colores[k], lty = ltipos[k])
}
legend("topright", legend = etiquetas,
       col = colores, lty = ltipos, lwd = 2, bty = "n")
```

A longer latent period delays the epidemic peak and produces a slightly flatter
curve, though the final number of cases barely changes because $R_0$ remains
constant.

## Incidence versus progression in the SEIR

In the SEIR model it is useful to distinguish two flows:

- **Incidence** ($\lambda S = \beta S I / N$): new exposures, flow $S \to E$.
- **Progression** ($\sigma E$): transition from exposed to infectious, flow $E \to I$.

During the epidemic growth phase incidence exceeds progression because individuals
are exposed faster than they become infectious.
Both curves cross around the peak and converge at the end.

```{r incidence-vs-progression}
# Access the derived variables stored in sim$derived
der <- sim_seir$derived

ylim_d <- max(der$incidence, der$progression) * 1.05

plot(der$time, der$incidence,
     type = "l", lwd = 2, col = "steelblue",
     ylim = c(0, ylim_d),
     xlab = "Time (days)", ylab = "Rate (individuals / day)",
     main = "SEIR: incidence (S->E) vs progression (E->I)")
lines(der$time, der$progression, lwd = 2, col = "firebrick", lty = 2)
legend("topright",
       legend = c("Incidence  (beta*S*I/N)", "Progression  (sigma*E)"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), lwd = 2, bty = "n")
```

The lag between both curves reflects exactly the latent period: newly exposed
individuals take on average $1/\sigma$ days to become infectious.

## Summary

| Feature                         | SIR              | SEIR             |
|---------------------------------|-----------------|-----------------|
| Compartments                    | S, I, R          | S, E, I, R       |
| Latent period                   | No               | Yes ($1/\sigma$) |
| $R_0 = \beta/\gamma$            | Yes              | Yes (same value) |
| Invasion threshold              | $R_0 > 1$        | $R_0 > 1$        |
| Final epidemic size             | Equal            | Equal            |
| Peak timing                     | Earlier          | Later            |

Using the SEIR instead of the SIR is especially relevant when the latent period
is long relative to the infectious period, as occurs with measles or COVID-19.
In those cases the SIR overestimates the initial growth rate and predicts the
peak too early.
