---
title: "From epidemic to endemic: SIR versus SIRS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From epidemic to endemic: SIR versus SIRS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4)
```

## Introduction

The key difference between the **SIR** and **SIRS** models is the duration of
immunity:

- **SIR**: permanent immunity. The pathogen generates a single wave and disappears.
- **SIRS**: temporary immunity. Recovered individuals become susceptible again at
  rate $\omega$, which allows **endemic** persistence and recurrent waves.

```{r cargar}
library(epiLabR)
```

## Shared parameters

```{r parametros}
beta  <- 0.3;  gamma <- 0.1   # transmission and recovery (day⁻¹)
R0    <- beta / gamma          # basic reproduction number = 3
N     <- 1e6;  I0 <- 20       # population and initial infectious individuals
cat("R0 =", R0, "| mean infectious period:", round(1 / gamma), "days\n")
```

With $R_0 = `r R0`$ the epidemic grows while there are enough susceptibles.

## 1. SIR model: single epidemic, permanent immunity

```{r sim-sir}
sim_sir <- simulate_epi(
  model = SIR_MODEL,
  times = 0:500,
  parms = c(beta = beta, gamma = gamma),
  init  = c(S = N - I0, I = I0, R = 0)
)
```

```{r plot-sir}
t     <- sim_sir$states$time
I_sir <- sim_sir$states$I

plot(t, I_sir,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Time (days)", ylab = "Infected I(t)",
     main = "SIR: single wave and extinction")
abline(h = 0, lty = 2, col = "grey60")
```

The SIR curve has a bell shape: it rises while $S > N/R_0$ and falls
when herd immunity exhausts the susceptibles. Eventually $I(t) \to 0$.

## 2. SIRS model: endemic state and recurrent waves

We add immunity loss at rate $\omega = 0.02$ day⁻¹, equivalent to a mean
immunity duration of $1/\omega = 50$ days.

```{r sim-sirs}
omega <- 0.02   # immunity loss rate (day⁻¹)

sim_sirs <- simulate_epi(
  model = SIRS_MODEL,
  times = 0:500,
  parms = c(beta = beta, gamma = gamma, omega = omega),
  init  = c(S = N - I0, I = I0, R = 0)
)

I_sirs <- sim_sirs$states$I
```

```{r plot-sir-vs-sirs}
ylim_max <- max(I_sir, I_sirs) * 1.05

plot(t, I_sir,
     type = "l", lwd = 2, col = "steelblue",
     ylim = c(0, ylim_max),
     xlab = "Time (days)", ylab = "Infected I(t)",
     main = "SIR vs SIRS: epidemic vs. endemic")
lines(t, I_sirs, lwd = 2, col = "firebrick")
legend("topright",
       legend = c("SIR (permanent immunity)",
                  "SIRS (omega = 0.02)"),
       col    = c("steelblue", "firebrick"),
       lwd    = 2, bty = "n")
```

The SIRS shows a first wave similar to the SIR, but the prevalence
oscillates and converges to a positive endemic level instead of going extinct.

## 3. Endemic equilibrium of the SIRS

For the SIRS model without demography, when $R_0 > 1$ a stable endemic
equilibrium exists. Setting the derivatives to zero yields:

$$S^* = \frac{N}{R_0}, \qquad
  I^* = N \cdot \frac{(\beta - \gamma)\,\omega}{\beta\,(\gamma + \omega)}$$

```{r equilibrio}
S_star <- N / R0
I_star <- N * (beta - gamma) * omega / (beta * (gamma + omega))
R_star <- N - S_star - I_star

cat("SIRS endemic equilibrium:\n")
cat("  S* =", round(S_star), "\n")
cat("  I* =", round(I_star), "\n")
cat("  R* =", round(R_star), "\n")

# Simulated value at the end of the simulation
I_final <- tail(I_sirs, 1)
cat("\nSimulated I(t=500):", round(I_final), "\n")
cat("Relative error:",
    round(abs(I_final - I_star) / I_star * 100, 2), "%\n")
```

```{r plot-equilibrio}
plot(t, I_sirs,
     type = "l", lwd = 2, col = "firebrick",
     xlab = "Time (days)", ylab = "Infected I(t)",
     main = "SIRS: convergence to the endemic equilibrium")
abline(h = I_star, lty = 2, lwd = 1.5, col = "darkred")
legend("topright",
       legend = c("Simulated I(t)",
                  paste0("I* = ", round(I_star))),
       col    = c("firebrick", "darkred"),
       lwd    = c(2, 1.5), lty = c(1, 2), bty = "n")
```

The simulated trajectory converges to the theoretical value $I^*$, confirming
that the endemic equilibrium predicts the long-run prevalence level.

## 4. Effect of the immunity loss rate (omega)

The larger $\omega$, the faster susceptibles are replenished, the higher the
endemic prevalence, and the shorter the intervals between waves.

```{r omega-comparacion}
omegas  <- c(omega_slow = 0.005, omega_medium = 0.02, omega_fast = 0.1)
colores <- c("seagreen", "firebrick", "darkorange")

# Theoretical I* for each omega
I_star_vec <- N * (beta - gamma) * omegas / (beta * (gamma + omegas))

# Simulate the three scenarios
sims_omega <- lapply(omegas, function(om) {
  simulate_epi(
    model = SIRS_MODEL,
    times = 0:500,
    parms = c(beta = beta, gamma = gamma, omega = om),
    init  = c(S = N - I0, I = I0, R = 0)
  )
})
```

```{r plot-omega}
ylim_om <- max(sapply(sims_omega, function(s) max(s$states$I))) * 1.05

plot(NA,
     xlim = c(0, 500), ylim = c(0, ylim_om),
     xlab = "Time (days)", ylab = "Infected I(t)",
     main = "Effect of omega on endemic dynamics")

for (k in seq_along(sims_omega)) {
  lines(sims_omega[[k]]$states$time,
        sims_omega[[k]]$states$I,
        lwd = 2, col = colores[k])
  abline(h = I_star_vec[k], lty = 2, col = colores[k])
}

legend("topright",
       legend = paste0("omega = ", omegas,
                       "  (I* ~ ", round(I_star_vec), ")"),
       col = colores, lwd = 2, bty = "n")
```

- With small $\omega$ (long-lasting immunity) the endemic level is very low.
- With large $\omega$ (short-lived immunity) susceptibles are replenished quickly:
  high endemic level and more frequent waves.
- In the limit $\omega \to 0$ the SIRS behaves like the SIR: extinction.

## Summary

| Feature                   | SIR                        | SIRS                              |
|---------------------------|----------------------------|-----------------------------------|
| Immunity                  | Permanent                  | Temporary (lasts $1/\omega$ days) |
| Long-run outcome          | Pathogen extinction        | Endemic state or recurrent cycles |
| Endemic equilibrium       | Does not exist             | $I^* > 0$ if $R_0 > 1$           |
| Real-world examples       | Measles (pre-vaccine)      | Flu, common cold, COVID-19        |
